\chapter{External protocols}

\section{Packages}

\subsection{Package for ordinary reader features}

The package for ordinary reader features is named
\texttt{eclector.reader}.  To use features of this package, we
recommend the use of explicit package prefixes, simply because this
package shadows and exports names that are also exported from the
\texttt{common-lisp} package.  Importing this package will likely
cause conflicts with the \texttt{common-lisp} package otherwise.

\subsection{Package for readtable features}

The package for readtable-related features is named
\texttt{eclector.readtable}.  To use features of this package, we
recommend the use of explicit package prefixes, simply because this
package shadows and exports names that are also exported from the
\texttt{common-lisp} package.  Importing this package will likely
cause conflicts with the \texttt{common-lisp} package otherwise.

\subsection{Package for CST features}

The package for features related to the creation of concrete syntax
trees is named \texttt{eclector.concrete-syntax-tree}.  Although this
package does not shadow any symbol in the \texttt{common-lisp}
package, we still recommend the use of explicit package prefixes to
refer to symbols in this package.

\section{Ordinary reader features}

\Defun {read} {\optional (input-stream \texttt{*standard-input*})\\
  (eof-error-p \texttt{t})
  (eof-value \texttt{nil})
  (recursive-p \texttt{nil})}

This function is the main entry point for the ordinary reader.  It is
entirely compatible with the standard \commonlisp{} function with the
same name.

\Defvar {*client*}

This variable is used by several generic functions called by
\texttt{read}.  The default value of the variable is \texttt{nil}.
Client code that wants to override or extend the default behavior of
some generic function of \sysname{} should bind this variable to some
standard object and provide a method on that generic function,
specialized to the class of that standard object.

\Defgeneric {read-common} {client input-stream eof-error-p eof-value}

This generic function is called by \texttt{read}, passing it the value
of the variable \texttt{*client*} and the corresponding parameters.
Client code can add methods on this function, specializing them to the
client class of its choice.  The actions that \texttt{read} need to
take for different values of the parameter \texttt{recursive-p} have
already been taken before \texttt{read} calls this generic function.

\Defgeneric {note-skipped-input} {client input-stream reason}

This generic function is called whenever the reader skips some input
such as a comment or a form that must be skipped because of a reader
conditional.  It is called with the value of the variable
\texttt{*client*}, the input stream from which the input is being read
and an object indicating the reason for skipping the input.  The
default method on this generic function does nothing.  Client code can
supply a method that specializes to the client class of its choice.

When this function is called, the stream is positioned immediately
\emph{after} the skipped input.  Client code that wants to know both
the beginning and the end of the skipped input must remember the
stream position before the call to \texttt{read} was made as well as
the stream position when the call to this function is made.

\Defvar {*skip-reason*}

This variable is used by the reader to determine why a range of input
characters has been skipped.  To this end, internal functions of the
reader as well as reader macros can set this variable to a suitable
value before skipping over some input.  Then, after the input has been
skipped, the generic function \texttt{note-skipped-input} is called
with the value of the variable as its \textit{reason} argument.

As an example, the method on \texttt{note-skipped-input} specialized
to\\
\texttt{eclector.concrete-syntax-tree:cst-client} relays the reason
and position information to the client by calling the\\
\texttt{eclector.concrete-syntax-tree:record-skipped-input} generic
function \seesec{sec:cst-reader-features}

\Defgeneric {read-token} {client input-stream eof-error-p eof-value}

This generic function is called by \texttt{read-common} when it has
been detected that a token should be read.  This function is
responsible for accumulating the characters of the token and then
calling \texttt{interpret-token} (see below) in order to create and
return a token.

\Defgeneric {interpret-token} {token token-escapes input-stream}

This generic function is called by \texttt{read-token} in order to
create a token from accumulated token characters.  The parameter
\textit{token} is a string containing the characters that make up the
token.  The parameter \textit{token-escapes} is a vector of Boolean
values (i.e. \texttt{nil} or \texttt{t}) with the same length as the
string \textit{token}.  It indicates, for each character in
\textit{token} whether the character was \emph{escaped} or read
normally.  This information is used to convert the characters in
\textit{token} according to the \emph{readtable case} of the current
readtable before a token is constructed.

\Defgeneric {interpret-symbol} {token position-package-marker-1\\
  position-package-marker-2 input-stream}

This generic function is called by the default method on
\texttt{interpret-token} when the syntax of the token corresponds to
that of a valid symbol.  The parameter \textit{token} is a string that
contains all the characters of the token.  The parameter
\textit{position-package-marker-1} contains the index into
\textit{token} of the first package marker, or \texttt{nil} if the
token contains no package markers.  The parameter
\textit{position-package-marker-2} contains the index into
\textit{token} of the second package marker, or \texttt{nil} if the
token contains no package markers or only a single package marker.
The parameter \textit{input-stream} is the input stream from which the
characters were read.  It can be used by client code to specialize the
behavior of this generic function.

\Defgeneric {call-reader-macro} {client input-stream char readtable}

This generic function is called when the reader has determined that
some character is associated with a reader macro.  The parameter
\textit{char} has to be used in conjunction with the
\textit{readtable} parameter to obtain the macro function that is
associated with the macro character.  The parameter
\textit{input-stream} is the input stream from which the reader macro
function will read additional input to accomplish its task.

The default method on this generic function simply obtain the reader
macro function for \textit{char} from \textit{readtable} and calls it,
passing \textit{input-stream} and \textit{char} as arguments.  The
default method therefore does the same thing that the standard
\commonlisp{} reader does.

\Defgeneric {evaluate-expression} {client expression}

This generic function is called by the default \texttt{\#.}~reader
macro function to perform read-time evaluation.  \textit{expression}
is the expression that should be evaluated as it was returned by a
recursive \texttt{read} call and potentially influenced by
\textit{client}.  The function has to either return the result of
evaluating \textit{expression} or signal an error.

The default method on this generic function simply returns the result
of \texttt{(cl:eval expression)}.

\Defgeneric {check-feature-expression} {client feature-expression}

This generic function is called by the default \texttt{\#+}- and
\texttt{\#-}-reader macro functions to check the well-formedness of
\textit{feature-expression} which has been read from the input stream
before evaluating it.  For compound expressions, only the outermost
expression is checked regarding the atom in operator position and its
shape -- child expressions are not checked.  The function returns an
unspecified value if \textit{feature-expression} is well-formed and
signals an error otherwise.

The default method on this generic function accepts standard
\commonlisp{} feature expression, i.e.\ expressions recursively
composed of symbols, \texttt{:not}-expressions,
\texttt{:and}-expressions and \texttt{:or}-expressions.

\Defgeneric {evaluate-feature-expression} {client feature-expression}

This generic function is called by the default \texttt{\#+}- and
\texttt{\#-}-reader macro functions to evaluate
\textit{feature-expression} which has been read from the input stream.
The function returns either true or false if
\textit{feature-expression} is well-formed and signals an error
otherwise.

For compound feature expressions, the well-formedness of child
expressions is not checked immediately but lazily, just before the
child expression in question is evaluated in a subsequent
\texttt{evaluate-feature-expression} call.  This allows feature
expression like \texttt{\#+(and my-cl-implementation (special-feature
  a b))} to succeed when the \texttt{:my-cl-implementation} feature is
absent.

The default method on this generic function first calls
\texttt{check-feature-expression} to check the well-formedness of
\textit{feature-expression}.  It then evaluates
\textit{feature-expression} according to standard \commonlisp{}
semantics for feature expressions.

\section{CST reader features}
\label{sec:cst-reader-features}

\Defun {cst-read} {\optional (input-stream \texttt{*standard-input*})\\
  (eof-error-p \texttt{t})
  (eof-value \texttt{nil})}

This function is the main entry point for the CST reader.  It is
mostly compatible with the standard \commonlisp{} function
\texttt{read}.  The differences are:

\begin{itemize}
\item The return value, unless \textit{eof-value} is returned, is an
  instance of a subclass of \texttt{concrete-syntax-tree:cst}.
\item The function does not accept a \textit{recursive} parameter
  since it sets up a dynamic environment in which calls to
  \texttt{eclector.reader:read} behave suitably.
\end{itemize}

\Defgeneric {make-source-range} {client start end}

This generic function is called in order to turn the range of input
characters designated by \textit{start} and \textit{end} into a range
representation suitable for \textit{client}.  The default method
returns \texttt{(cons start end)}.

\Defgeneric {record-skipped-input} {client input-stream reason source}

This generic function is called after the reader skipped over a range
of characters in \textit{input-stream}.  The value of the
\textit{source} parameter delimits the skipped range using start and
end positions obtained via \texttt{source-position}.

Reasons for skipping input include comments, the \texttt{\#+} and
\texttt{\#-} reader macros and \texttt{*read-suppress*}.  The
aforementioned reasons are reflected by the value of the
\textit{reason} parameter as follows:

\begin{tabular}{ll}
  Input                                          & Value of the \textit{reason} parameter\\
  \hline
  Comment starting with $n$ \texttt{;}           & \texttt{(:line-comment . $n$)}\\
  Comment delimited by \texttt{\#|} \texttt{|\#} & \texttt{:block-comment}\\
  \texttt{\#+\textit{false-feature-expression}}  & \texttt{:reader-macro}\\
  \texttt{\#-\textit{true-feature-expression}}   & \texttt{:reader-macro}\\
  \texttt{*read-suppress*} is true               & \texttt{*read-suppress*}
\end{tabular}

The default method does nothing.
